---
output: github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# Exploratory Data Analysis with R/RStudio

[![Travis-CI Build Status](https://travis-ci.org/rnabioco/eda.svg?branch=master)](https://travis-ci.org/rnabioco/eda)

## Syllabus

This short course teaches you to analyze and visualize large, complex data sets use the R statistical programming language, RStudio, and new tools available in the "tidyverse". We use common biological experiments to illustrate analysis strategies including studies of mRNA expression, histone post-translational modification in the context of chromatin, and cell-type characterization using single cell approaches.

### Course number

IDPT 7810 006

### Dates and Times

The course (2 credit hours) consists of seven 2 hour classes held Mon, Wed and Fri from Nov 29 through Dec 15 in the Health Sciences Library Teaching Lab 2 from 1:00-3:00 pm.

### Instructor

Jay Hesselberth (jay.hesselberth@gmail.com)

### Co-instructors

Austin Gillen (austin.gillen@gmail.com)

Kent Riemondy (kent.riemondy@gmail.com) 

## Course marks

Your grade will be based on several factors:

- Quizzes (50%)
- Exercises (25%) 
- Attendance (25%)

Quizzes (25 points) are assigned at the end of each class. We will try to leave 15 minutes at the end so that you can preview the quiz and ask questions. 

Exercises (50 points) are assigned on Fridays. In these exercises you pose a specific hypothesis and then write up your analysis that supports or refutes the hypothesis. We included an example to give you a framework.

## Installation

You can install `eda` from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("rnabioco/eda")
```
