---
output: github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# Exploratory Data Analysis with R/RStudio

[![Travis-CI Build Status](https://travis-ci.org/jayhesselberth/eda.svg?branch=master)](https://travis-ci.org/jayhesselberth/eda)

## Syllabus

This short course teaches you to analyze and visualize large, complex data sets use the R statistical programming language, RStudio, and new tools available in the "tidyverse". We use common biological experiments to illustrate analysis strategies including studies of mRNA expression, histone post-translational modification in the context of chromatin, and cell-type characterization using single cell approaches.

Completion of this course is a pre-requisite for taking the [Genome Analysis Workshop](http://molb7621.github.io/workshop/), a popular course taught in the Spring semester.

### Dates and Times

The course (2 credit hours) consists of eight 2 hour classes held Mon, Wed and Fri from Nov 29 through Dec 15 in the Health Sciences Library Teaching Lab 2 from 1:00-3:00 pm.

### Instructor

Jay Hesselberth (jay.hesselberth@gmail.com)

### Co-instructors

Austin Gillen (austin.gillen@gmail.com)

Kent Riemondy (kent.riemondy@gmail.com) 

## Course marks

Your grade will be based on several factors:

- Attendance (25%)
- Quizzes (25%)
- Exercises (50%) 

Quizzes are assigned at the end of each class. We will try to leave 15 minutes at the end so that you can preview the quiz and ask questions.

Exercises are assigned on Friday. These are 

## Installation

You can install eda from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("jayhesselberth/eda")
```
