---
output: github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# Exploratory Data Analysis with R/RStudio

[![Travis-CI Build Status](https://travis-ci.org/rnabioco/eda.svg?branch=master)](https://travis-ci.org/rnabioco/eda)

## Syllabus

This short course teaches you to analyze and visualize large, complex data sets use the R statistical programming language, RStudio, and new tools available in the "tidyverse". We use common biological experiments to illustrate analysis strategies including studies of mRNA expression, histone post-translational modification in the context of chromatin, and cell-type characterization using single cell approaches.

### Course number

IDPT 7810 006

### Dates and Times

The course (2 credit hours) consists of seven 2 hour classes held Mon, Wed and Fri from Nov 29 through Dec 15 in the Health Sciences Library Teaching Lab 2 from 1:00-3:00 pm.

We have dedicated office hours on Thurs from 1-2:30pm in RC1 South 9th floor in the main office (behind the glass walls). Come with questions, problems, etc.

### Instructor

Jay Hesselberth (jay.hesselberth@gmail.com)

### Co-instructors

Austin Gillen (austin.gillen@gmail.com)

Kent Riemondy (kent.riemondy@gmail.com) 

## Course marks

See [Grading and Assignments](https://rnabioco.github.io/eda/articles/assignments.html).

## Installation

You can install `eda` from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("rnabioco/eda")
```
