---
output: github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# Practical Biological Data Analysis with R/RStudio

## Syllabus

This short course teaches you to analyze and visualize large, complex biological data sets using the R statistical programming language, RStudio, and new tools available in the "tidyverse". We will focus on key analysis skills and foundational programming concepts necessary for the efficient and reproducible analysis of biological data sets.

### Course number

IDPT 7810 006

### Dates and Times

The course (2 credit hours) consists of 12 2 hour classes beginning on Nov 25 and 26, and continuing Mon through Fri from December 2 through December 13 in P18-CTL-1309 (RC1 North). Classes will be held 1:00-3:00 pm. 

We have dedicated office hours on Thurs from 2:30-4pm in RC1 South 9th floor in the main office (behind the glass walls). Come with questions, problems, etc. Please send an email to `rbi.fellows@cuanschutz.edu` to let us know you're coming.

### Instructor

Kent Riemondy (kent.riemondy@cuanschutz.edu)

### Co-instructors

Ryan Sheridan (ryan.sheridan@cuanschutz.edu)

## Course marks

See [Grading and Assignments](https://rnabioco.github.io/practical-data-analysis/articles/assignments.html).

## Installation

You can install the course package from github with:

```{r gh-installation, eval = FALSE}
# install.packages("pak")
pak::pkg_install("rnabioco/practical-data-analysis")
```
