---
title: "Assignments and Grading"
author: "Jay Hesselberth"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Grading and Assignments}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "img/class-1-",
  fig.align = "center"
)
```

# Grades

Your grade is based on a combination of:

- Pre-requisite completion (5%)
- Attendance and participation (15%)
- Assignments (80%)

# Assignments

There are two types of assignments in the course:

- **Nightly quizzes** worth 30 points each.
- **Weekend analyses** worth 60 points each.

Assignments must be turned in as Rmarkdown text files via Canvas. **They must render completely when the Knit button is clicked**. This means you will need to load all libraries within the document. This is a good chunk to put in your Rmarkdown code:

```{r eval=FALSE}
library(tidyverse)
library(eda)
```

You are welcome and encouraged to work in groups to complete the assignments. One stipulation: if you work together, tell us who your collaborators were in the assignment.

# Grading Rubric

Each assignment problem will be evaluated using the following rubric:

1. Text annotations are provided laying out the problem, the strategy, and the conclusions.
2. Code is succinct and commented.
3. Graph types and layouts are chosen carefully to illustrate a single point.

Other required features include:

- **Remove extraneous information from your submission.** If you start with the template from `File > New File > R Markdown...`, then delete the generic sections.

- **Include the text of the question into your submission.** Use markdown to delineate the Question, Strategy and Interpretation sections. Here is an example layout.

```
# Question 1

Assess the relationship between hp and cyl in `mtcars`.

## Strategy

Plot hp vs cyl and examine.

## Interpretation

There is a linear relationship between hp and cyl.
```

- Add [`code_folding`](rmarkdown.rstudio.com/html_document_format.html#code_folding) to your front matter so that code appears next to plots etc.

