---
title: "Assignments and Grading"
author: "Jay Hesselberth"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Grading and Assignments}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "img/class-1-",
  fig.align = "center"
)
```

# Grades

Your grade is based on a combination of:

- Pre-requisite completion (5%)
- Attendance and Participation (15%)
- Assignments (80%)

# Assignments

There are two types of assignments in the course:

- Nightly quizzes, where questions are worth 5 points each.
- Weekend analyses, where questions are worth 10 points each.

Assignments must be turned in as Rmarkdown text files via Canvas. **They must render completely when the Knit button is clicked**. This menas you will need to load all libraries within the document. This is a good chunk to put in your Rmarkdown code:

```{r eval=FALSE}
library(tidyverse)
library(eda)
```

You are welcome and encouraged to work in groups to complete the assignments. One stipulation: if you work together, tell us who your collaborators were in the assignment.

# Grading Rubric

Each assignment problem will be evaluated using the following rubric:

- Text annotations are provided laying out the problem, the strategy, and the conclusions.
- Code is succinct and commented.
- Graph types and layouts are chosen carefully to illustrate a single point.
