---
title: "Assignments and Grading"
author: "Jay Hesselberth"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Grading and Assignments}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "img/class-1-",
  fig.align = "center"
)
```

# Grades

Your grade is based on a combination of:

- Pre-requisite completion (5%)
- Attendance and participation (5%)
- Assignments (90%)

# Assignments

There are two types of assignments in the course:

- **Nightly quizzes** worth 60 points each.
- **Weekend analyses**, also worth 60 points each.

Assignments must be turned in via Github. **They must render completely when the Knit button is clicked**. This means you will need to load all libraries within the document.

This is a good way to start your Rmarkdown document:

```{r eval=FALSE}
library(tidyverse)
library(pbda)
```

You are welcome and encouraged to work in groups to complete the assignments, with one stipulation: if you work together, tell us who your collaborators were in the assignment.

# Grading Rubric

Each assignment problem will be evaluated using the following rubric:

| **Topic** | **Excellent (10)** | **Satisfactory (8)** | **Needs work (6)** |
| --------- | ------------- | ---------------- | -------------- |
| Coding Style | ✔️           |                  |                |
| Coding strategy |  ✔️          |                  |                |
| Presentation: graphs |   ✔️         |                  |                |
| Presentation: tables |  ✔️          |                  |                |
| Achievement, creativity |   ✔️         |                  |                |
| Ease of access | ✔️           |                 |                |

**Grade: (60 points)**

Before you submit, please ensure you also do the following:

- **Remove extraneous information from your submission.** If you start with the template from `File > New File > R Markdown...`, then delete the generic sections.

- **Include the text of the question into your submission.** Use markdown to delineate the Question, Strategy and Interpretation sections. Here is an example layout.

```
# Question 1

Assess the relationship between hp and cyl in `mtcars`.

## Strategy

Plot hp vs cyl and examine.

## Interpretation

There is a linear relationship between hp and cyl.
```

- Add [`code_folding`](rmarkdown.rstudio.com/html_document_format.html#code_folding) to your front matter so that code appears next to plots etc.

