---
title: "Class 7: Iteration"
author: "Ryan Sheridan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Class-7}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = F}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "left"
)
library(pbda)
library(tidyverse)
library(cowplot)
```

<br>

### Introduction

As discussed in the previous class, you should try to limit duplication in your code. One way to do this is by writing functions, another way is through iteration. Reducing code duplication has several benefits:

* Your code easier to read
* You reduce the chance of mistakes from repeated copying and pasting
* It is easier to adapt your code for different requirements

Further reading

* [R for Data Science](https://r4ds.had.co.nz) by Garrett Grolemund and Hadley Wickham
* [Advanced R](https://adv-r.hadley.nz) by Hadley Wickham

<br>

<br>

### `for` loops
`for` loops allow you to run the same code block repeatedly without copying and pasting.
```{r}
# When writing a for loop, think about the output, sequence, and body
out <- vector("numeric", 5)  # output

for (i in 1:length(out)) {   # sequence
  
  out[i] <- i * 5            # body
  
}

# It's helpful to think about what happens during each cycle of the loop
out[1] <- 1 * 5  # i = 1
out[2] <- 2 * 5  # i = 2
out[3] <- 3 * 5  # i = 3
out[4] <- 4 * 5  # i = 4
out[5] <- 5 * 5  # i = 5
```

```{r}
# Lets use rnorm() to create 5 vectors with different values for 'mean'
# One way to do this is by copying and pasting
vec_in <- c(1, 50, 20, 5, 70)                # input

out <- vector("list", length(vec_in))        # output

out[[1]] <- rnorm(n = 10, mean = vec_in[1])    
out[[2]] <- rnorm(n = 10, mean = vec_in[2])
out[[3]] <- rnorm(n = 10, mean = vec_in[3])
out[[4]] <- rnorm(n = 10, mean = vec_in[4])
out[[5]] <- rnorm(n = 10, mean = vec_in[5])

# Use a for loop to reduce code duplication
vec_in <- c(1, 50, 20, 5, 70)                  # input

out <- vector("list", length(vec_in))          # output

for (i in 1:length(vec_in)) {                  # sequence
  
  out[[i]] <- rnorm(n = 10, mean = vec_in[i])  # body
  
}
```

<br>

*Write a `for` loop to create 3 vectors of different lengths*
```{r, echo = F}
vec_in <- c(5, 10, 2)                  # input

out <- vector("list", length(vec_in))  # output

for (i in 1:length(vec_in)) {          # sequence
  
  out[[i]] <- rnorm(n = vec_in[i])     # body
  
}

out
```

<br>

Using `seq_along(x)` instead of `1:length(x)`
```{r}
# seq_along() mimics 1:length() for non-empty vectors
vec_in <- c(5, 10, 2)

1:length(vec_in)

seq_along(vec_in)

# seq_along() correctly handles empty vectors
emp_vec <- vector("numeric", 0)

1:length(emp_vec)

seq_along(emp_vec)
```

<br>

<br>

### Lets try this with the Brauer data
```{r, fig.width = 10, fig.height = 8}
# We want to create a supplemental figure showing the growth rate vs expression for four genes
plot_expr <- function(input, sys_name, ...) {
  
  gg_data <- input %>%
    filter(systematic_name == sys_name)
  
  plot_title <- gg_data$name[1]
  plot_sub <- gg_data$MF[1]
  
  gg_data %>%
    ggplot(aes(rate, expression, color = nutrient)) +
    geom_point(...) +
    labs(title = plot_title, subtitle = plot_sub) +
    facet_wrap(~ nutrient) +
    theme_cowplot() +
    theme(legend.position = "none")
}

# Lets try this with the copy-and-paste method
vec_in <- c("YDL104C", "YLR115W", "YMR183C", "YML017W")       # input

out <- vector("list", length(vec_in))                         # output 

out[[1]] <- plot_expr(brauer_gene_exp, sys_name = vec_in[1])  
out[[2]] <- plot_expr(brauer_gene_exp, sys_name = vec_in[2])
out[[3]] <- plot_expr(brauer_gene_exp, sys_name = vec_in[3])
out[[4]] <- plot_expr(brauer_gene_exp, sys_name = vec_in[4])

plot_grid(plotlist = out)
```

<br>

*Re-write the code from above using a `for` loop to generate our figure*
```{r, eval = F}
vec_in <- c("YDL104C", "YLR115W", "YMR183C", "YML017W")  # input

out <- vector("list", length(vec_in))                    # output

for ( ???? ) {                                           # sequence
  
                                                         # body

}

plot_grid(plotlist = out)
```

<br>

<br>

### Introduction to `map()`
`for` loops are a powerful tool to reduce code duplication, however your code can be further simplified using the `map()` functions provided in the purrr package. These `map()` functions essentially run `for (i in seq_along(x))` behind the scenes so you don't have to explicitly type this.

There is a function for each type of output:

- `map()` makes a list
- `map_lgl()` makes a logical vector
- `map_int()` makes an integer vector
- `map_dbl()` makes a double vector
- `map_chr()` makes a character vector

Each `map()` function requires two inputs: `map(.x, .f, ...)`

- `.x` is a list or atomic vector
- `.f` is a function or formula
```{r}
# We previously used a for loop to create vectors with different values for mean
vals <- c(1, 50, 20, 5, 70)                  # input

out <- vector("list", length(vals))          # output

for (i in seq_along(vals)) {                 # sequence
  
  out[[i]] <- rnorm(n = 10, mean = vals[i])  # body
  
}

# Using map() we can further simplify this code
# .x indicates where each element of the vector should be inserted
out <- map(
  .x = vals,
  .f = ~ rnorm(n = 10, mean = .x)
)

# You can use brackets to include a multi-line code block
out <- map(vals, ~ {
  
  rnorm(n = 10, mean = .x)

})

# map() allows for very readable code
# Each element of the vector is passed to the first available argument
out <- map(vals, rnorm, n = 10)
```

<br>

*Use `rnorm()` and `map()` to create 3 vectors of different lengths*
```{r, echo = F}
lens <- c(5, 10, 2)

out <- map(lens, ~ rnorm(n = .x))

out <- map(lens, rnorm)

out
```

<br>

*Re-write the code from above using `map()` to generate our growth rate figure*
```{r, eval = F}
genes <- c("YDL104C", "YOR069W", "YLR115W", "YPR036W")

expr_plots <- map(
  
  ????
  
)

plot_grid(plotlist = expr_plots)
```

```{r, fig.width = 10, fig.height = 8, echo = F}
genes <- c("YDL104C", "YOR069W", "YLR115W", "YPR036W")

expr_plots <- map(
  .x = genes, 
  .f = plot_expr, 
  input = brauer_gene_exp
)

plot_grid(plotlist = expr_plots)
```

<br>

<br>

### Mapping over multiple arguments
If you have two vectors containing values that you want to pass to a function this can be accomplished with `map2()`.
```{r, fig.width = 10, fig.height = 8}
genes <- c("YDL104C", "YOR069W", "YLR115W", "YPR036W")
shapes <- c(1, 2, 3, 4)

expr_plots <- map2(genes, shapes, ~ {
  plot_expr(
    input    = brauer_gene_exp,
    sys_name = .x,
    shape    = .y
  )
})

plot_grid(plotlist = expr_plots)
```

<br>

*Use `map2()` to create plots for 4 different genes, each with a different point size*
```{r, eval = F}
genes <- c("YDL104C", "YOR069W", "YLR115W", "YPR036W")

expr_plots <- map2(
  
  ????
  
)

plot_grid(plotlist = expr_plots)
```

```{r, fig.width = 10, fig.height = 8, echo = F}
genes <- c("YDL104C", "YOR069W", "YLR115W", "YPR036W")
sizes <- c(1, 2, 4, 8)

expr_plots <- map2(genes, sizes, ~ {
  plot_expr(
    input    = brauer_gene_exp,
    sys_name = .x,
    size     = .y
  )
})

plot_grid(plotlist = expr_plots)
```

<br>

`pmap` can be used to map over any number of arguments.
```{r, fig.width = 10, fig.height = 8}
# Create a list of input vectors
genes <- c("YDL104C", "YOR069W", "YLR115W", "YPR036W")
sizes <- c(2, 3, 4, 5)
shapes <- c(1, 2, 3, 4)

plot_args <- list(genes, sizes, shapes)

# Use an argument list with pmap
expr_plots <- pmap(plot_args, ~ {
  plot_expr(
    input = brauer_gene_exp,
    sys_name = ..1,
    size     = ..2,
    shape    = ..3
  )
})

# A simpler way
plot_args <- list(
  sys_name = c("YDL104C", "YOR069W", "YLR115W", "YPR036W"),
  size     = c(2, 3, 4, 5),
  shape    = c(1, 2, 3, 4)
)

expr_plots <- pmap(
  .l = plot_args, 
  .f = plot_expr, 
  input = brauer_gene_exp
)

plot_grid(plotlist = expr_plots)
```



