---
title: "Class 6: Writing Functions"
author: "Ryan Sheridan"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = F}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "left"
)
library(pbda)
library(tidyverse)
library(cowplot)
```

<br>

## What is a function?
As an analyst you will eventually find yourself in the position of wanting to reuse a block of code. There are two general ways to do this:

1. copy-and-paste
2. write a function

A function is essentially a block of code that you've given a name and saved for later. Functions have several advantages:

* They make your code easier to read
* They reduce the chance of mistakes from repeated copying and pasting
* They make it easier to adapt your code for different requirements

Further reading

* [R for Data Science](https://r4ds.had.co.nz) by Garrett Grolemund and Hadley Wickham
* [Advanced R](https://adv-r.hadley.nz) by Hadley Wickham
```{r, eval = F}
library(pbda)
library(tidyverse)
library(cowplot)
```

```{r}
# An example: you want to rescale a numeric vector so all values are between 0 and 1
a <- rnorm(n = 10)
a

rng <- range(a)
(a - rng[1]) / (rng[2] - rng[1])

# What if we want to repeat this on other vectors?
# One way is to copy and paste
b <- rnorm(n = 10)
c <- rnorm(n = 10)

rng <- range(b)
new_b <- (b - rng[1]) / (rng[2] - rng[1])

rng <- range(c)
new_c <- (c - rng[1]) / (rng[2] - rng[1])

# A better way is to write a function...
```

<br>

<br>

## How to write a function
There are three general steps for writing functions:

1. Pick a name
2. Identify the inputs
3. Add code to the body
```{r, eval = F}
# Lets write a function to rescale a numeric vector
rescale_vec <- function(x) {
  
  rng <- range(x)
  (x - rng[1]) / (rng[2] - rng[1])

}

rescale_vec(b)
rescale_vec(c)
```

<br>

*Write functions for the following bits of code*
```{r, eval = F}
# function 1
x / sum(x)

# function 2
(x + y) / z

# function 3
sqrt(sum((x - mean(x))^2) / (length(x) - 1))
```

```{r, echo = F}
calc_sd <- function(x) {
  sqrt(sum((x - mean(x))^2) / (length(x) - 1))
}

calc_sd <- function(x) {
  l <- length(x) - 1
  m <- mean(x)
  v <- sum((x - m)^2) / l
  sqrt(v)
}
```

<br>

<br>

## The function execution environment
- When running a function an execution environment is created, which is separate from the global environment
- The execution environment contains objects created within the function
- The execution environment follows the "fresh start" principle
- When R searches for an object referenced by a function, the execution environment takes precedence

<br>

*Can objects present in the global environment be referenced from within a function?*
```{r}
# Earlier we saved a numeric vector "a"
a

sum_nums <- function(x) {
  x + a
}

# Yes!
sum_nums(10)
```

<br>

*Can code executed within a function modify an object present in the global environment?*
```{r}
sum_nums <- function(x) {
  a <- x + a
}

# When we run sum_nums(), will this overwrite our original vector?
sum_nums(10)

# No! (not when using the '<-' assignment operator)
a
```

<br>

<br>

## A more relevant example
*Using the Brauer data lets create a scatter plot comparing growth rate vs expression for the gene YDL104C. Use `facet_wrap()` to create a separate plot for each nutrient.*
```{r, fig.width = 6, fig.height = 4.5, echo = F}
gg_data <- brauer_gene_exp %>%
  filter(systematic_name == "YDL104C")               # Filter for gene of interest

gg_data %>%
  ggplot(aes(rate, expression, color = nutrient)) +  # Create scatter plot
  geom_point(size = 2) +
  facet_wrap(~ nutrient) +                           # Create separate plot for each nutrient
  theme_cowplot() +
  theme(legend.position = "none")
```

<br>

What if you want to create this plot for other genes? *Write a function the takes a data.frame and systematic_name as inputs and creates scatter plots for each nutrient*
```{r, eval = F}
# Fill in the function body
# You can include default values for your arguments
plot_expr <- function(input, sys_name = "YNL049C") {
  
  ????
  
}
```

```{r, echo = F}
plot_expr <- function(input, sys_name = "YNL049C") {
  gg_data <- input %>%
    filter(systematic_name == sys_name)
  
  gg_data %>%
    ggplot(aes(rate, expression, color = nutrient)) +
    geom_point(size = 2) +
    facet_wrap(~ nutrient) +
    theme_cowplot() +
    theme(legend.position = "none")
}
```

```{r, fig.width = 6, fig.height = 4.5}
p <- plot_expr(
  input = brauer_gene_exp, 
  sys_name = "YDL104C"
)

# You can also use the %>% pipe with your custom functions
p <- brauer_gene_exp %>%
  plot_expr(sys_name = "YDL104C")

p
```

<br>

*Modify our plotting function to add the gene name as the plot title and the molecular function (MF) as a subtitle*
```{r, echo = F}
plot_expr <- function(input, sys_name) {
  gg_data <- input %>%
    filter(systematic_name == sys_name)
  
  plot_title <- gg_data$name[1]
  plot_sub <- gg_data$MF[1]
  
  gg_data %>%
    ggplot(aes(rate, expression, color = nutrient)) +
    geom_point(size = 2) +
    labs(title = plot_title, subtitle = plot_sub) +
    ggtitle(plot_title) +
    facet_wrap(~ nutrient) +
    theme_cowplot() +
    theme(legend.position = "none")
}
```

```{r, fig.width = 6, fig.height = 4.5}
brauer_gene_exp %>%
  plot_expr("YDL104C")
```

<br>

<br>

## Conditional statements
`if` statements allow you to execute code depending on defined conditions.
```{r, eval = F}
if (condition) {
  code executed when condition is TRUE
  
} else {
  code executed when condition is FALSE
}
```

R has a set of operators that can be used to write conditional statements

Operator | Description
:-------:|:-----------
<        | less than
<=       | less or equal
\>       | greater than
\>=      | greater or equal
==       | equal
!=       | not equal
!x       | not x
x \|\| y | x or y
x && y   | x and y
x %in% y | x is present in y

<br>

*Add an `if` statement to our plotting function to account for a missing gene name*
```{r, eval = F}
plot_expr <- function(input, sys_name) {
  gg_data <- input %>%
    filter(systematic_name == sys_name)
  
  plot_title <- gg_data$name[1]
  plot_sub <- gg_data$MF[1]
  
  ????
  
  gg_data %>%
    ggplot(aes(rate, expression, color = nutrient)) +
    geom_point(size = 2) +
    labs(title = plot_title, subtitle = plot_sub) +
    facet_wrap(~ nutrient) +
    theme_cowplot() +
    theme(legend.position = "none")
}
```

```{r, echo = F}
plot_expr <- function(input, sys_name) {
  gg_data <- input %>%
    filter(systematic_name == sys_name)
  
  plot_title <- gg_data$name[1]
  plot_sub <- gg_data$MF[1]
  
  if (plot_title == "") {
    plot_title <- sys_name
  }
  
  gg_data %>%
    ggplot(aes(rate, expression, color = nutrient)) +
    geom_point(size = 2) +
    labs(title = plot_title, subtitle = plot_sub) +
    facet_wrap(~ nutrient) +
    theme_cowplot() +
    theme(legend.position = "none")
}
```

```{r, fig.width = 6, fig.height = 4.5}
brauer_gene_exp %>%
  plot_expr("YNL095C")
```

<br>

Conditional statements can be linked together
```{r, eval = F}
# Using 'else if'
if (condition_1) {
  executed when condition_1 is TRUE
  
} else if (condition_2) {
  executed when condition_1 is FALSE and condition_2 is TRUE
  
} else {
  executed when condition_1 and condition_2 are FALSE
}

# The 'and' operator
if (condition_1 && condition_2) {
  executed when condition_1 and condition_2 are TRUE
  
} else {
  executed when condition_1 or condition_2 are FALSE
}

# The 'or' operator
if (condition_1 || condition_2) {
  executed when condition_1 or condition_2 are TRUE
  
} else {
  executed when condition_1 and condition_2 are FALSE
}
```

<br>

<br>

## Checking inputs
When writing functions it can be useful to check input values to make sure they are valid. Lets modify our plotting function to check that `sys_name` is a string.

- `is.character()`
- `is.numeric()`
- `is.logical()`
- `is.factor()`

```{r, fig.width = 6, fig.height = 4.5}
plot_expr <- function(input, sys_name) {
  
  if (!is.character(sys_name)) {
    stop("sys_name must be a string!")
  }
  
  gg_data <- input %>%
    filter(systematic_name == sys_name)
  
  plot_title <- gg_data$name[1]
  plot_sub <- gg_data$MF[1]
  
  if (plot_title == "") {
    plot_title <- sys_name
  }
  
  gg_data %>%
    ggplot(aes(rate, expression, color = nutrient)) +
    geom_point(size = 2) +
    labs(title = plot_title, subtitle = plot_sub) +
    facet_wrap(~ nutrient) +
    theme_cowplot() +
    theme(legend.position = "none")
}

brauer_gene_exp %>%
  plot_expr("YDL104C")
```

<br>

*Modify our plotting function to check that `sys_name` is present in the input. Hint: try the `%in%` operator*
```{r, eval = F}
plot_expr <- function(input, sys_name) {
  
  if (!is.character(sys_name)) {
    stop("sys_name must be a string!")
  }
  
  if ( ???? ) {
    stop( ???? )
  }
  
  gg_data <- input %>%
    filter(systematic_name == sys_name)
  
  plot_title <- gg_data$name[1]
  plot_sub <- gg_data$MF[1]
  
  if (plot_title == "") {
    plot_title <- sys_name
  }
  
  gg_data %>%
    ggplot(aes(rate, expression, color = nutrient)) +
    geom_point(size = 2) +
    labs(title = plot_title, subtitle = plot_sub) +
    facet_wrap(~ nutrient) +
    theme_cowplot() +
    theme(legend.position = "none")
}
```

```{r, echo = F}
plot_expr <- function(input, sys_name) {
  
  if (!is.character(sys_name)) {
    stop("sys_name must be a string!")
  }
  
  if (!sys_name %in% input$systematic_name) {
    stop("sys_name not found in input data!")
  }
  
  gg_data <- input %>%
    filter(systematic_name == sys_name)
  
  plot_title <- gg_data$name[1]
  plot_sub <- gg_data$MF[1]
  
  if (plot_title == "") {
    plot_title <- sys_name
  }
  
  gg_data %>%
    ggplot(aes(rate, expression, color = nutrient)) +
    geom_point(size = 2) +
    labs(title = plot_title, subtitle = plot_sub) +
    facet_wrap(~ nutrient) +
    theme_cowplot() +
    theme(legend.position = "none")
}
```

<br>

<br>

## Passing arguments with the ellipsis (...)
The ellipsis allows a function to take an arbitrary number of arguments, which can then be passed to an inner function. This is nice when you have an inner function that has a lot of useful arguments. Lets first try this with our simple `rescale_vec()` function. 
```{r}
rescale_vec <- function(x, ...) {
  rng <- range(x, ...)
  (x - rng[1]) / (rng[2] - rng[1])
}

rescale_vec(a)

a[1] <- NA

rescale_vec(a, na.rm = T)
```

<br>

*Modify our plotting function so the user can change the point size, shape, and alpha*
```{r, fig.width = 6, fig.height = 4.5}
# A cumbersome way
plot_expr <- function(input, sys_name, pt_size = 2, pt_shape = 1, pt_alpha = 1) {
  input %>%
    filter(systematic_name == sys_name) %>%
    ggplot(aes(rate, expression, color = nutrient)) +
    geom_point(size = pt_size, shape = pt_shape, alpha = pt_alpha) +
    facet_wrap(~ nutrient) +
    theme_cowplot() +
    theme(legend.position = "none")
}

# With the ellipsis
plot_expr <- function(input, sys_name, ...) {
  input %>%
    filter(systematic_name == sys_name) %>%
    ggplot(aes(rate, expression, color = nutrient)) +
    geom_point(...) +
    facet_wrap(~ nutrient) +
    theme_cowplot() +
    theme(legend.position = "none")
}

# Now we can easily change the point size and shape
plot_expr(
  input = brauer_gene_exp,
  sys_name = "YDL104C",
  size = 5,
  shape = 2,
  alpha = 0.75
)
```

<br>

<br>

## Saving your functions for later
A good way to save commonly used functions is to keep them in a separate R script. You can load your functions using the `source()` command.
```{r, eval = F}
source("path/to/my_functions.R")
```
