---
title: "Class 5: Basic R data types and structures"
author: "Rui Fu"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Class-5}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "#>",
  fig.align = "center"
)
```

## Before we get started: update pbda

```{r}
#devtools::install_github("rnabioco/practical-data-analysis")
#pbda::update_pbda() # probably say no to updating other packages
library(pbda)
library(dplyr)
```

## Classes to come:

1. basic programming concepts (functions, conditional statements, for loops, etc)

2. RNAseq

3. gene lists, GO term enrichment

4. other tips and tricks

# atom (single value) data types

1. character

2. integer

3. numeric

4. logical

### look at classes
```{r}
class(0L)
class(0.0)
class("0")

is.numeric(0L) # test for certain type/class
is.integer(0L)
```

### coerce to other classes
```{r, type1}
class(as.character(0))
class(as.numeric("0"))
```

### integer vs numeric
```{r, type2}
1:5
class(1:5)
object.size(1:1000)
object.size(as.numeric(1:1000)) # integer saves space
```

### be careful with integer coercion
```{r, type3}
as.integer(1.8) # note that as.integer isn't rounding
round(1.8)

as.integer(-1.8) # be very careful
round(-1.8)
```

### numbers to logical
```{r, type4}
as.logical(0) # 0 == FALSE
as.logical(0.1) # anything not 0 is TRUE
as.logical("0.1") # can't coerce
```

# data structures

## 1. atomic vector, a combination of values

### indexing
```{r, vector1}
v1 <- c("geneA", "geneB", "geneC")
length(v1)

v1[2] # access element by index
v1[c(3, 1)] # use index to change order
v1[c(TRUE, TRUE, FALSE)] # use logical vector as index

v2 <- c(5, 10, 0) # pretend that v2 contains expression values for v1, they can be filtered like this:
v2 >= 2 # result is logical vector
v1[v2 >= 2]
```

### combining and coercion
```{r, vector2}
c(v2, 4) # combine values
c(4, v2) # note order
c(v1, v2) # vectors only contain same type of data

class(v2)
as.character(v2) # coercion over entire vector
as.character(v2) %>% class()
```

### useful functions
```{r, vector3}
unique(c(1, 2, 3, 2))

sort(c(2, 4, 3))
sort(c("geneB", "geneA", "geneC"))
sort(c("geneB", "geneA", "geneC"), decreasing = TRUE)

intersect(c(1, 2, 3), c(2, 3, 4))

setdiff(c(1, 2, 3), c(2, 3, 4))
setdiff(c(2, 3, 4), c(1, 2, 3)) # note difference order makes
```

## 2. data.frame, combination of multiple vectors

### df vs tbl
```{r, df1}
mtcars
class(mtcars)

mtcars_tbl <- as_tibble(mtcars, rownames = "name") # tibble usually drops rownames
mtcars_tbl # only prints first 10 by default
class(mtcars_tbl) # still a data.frame, but more
class(mtcars_tbl) == "data.frame" # programming without considering potentially different result structure is dangerous
is.data.frame(mtcars_tbl)  
```

### exploring data.frame
```{r, df2}
mtcars$mpg # a vector
mtcars[["mpg"]] # also a vector
dim(mtcars)
ncol(mtcars)
nrow(mtcars)
colnames(mtcars)
rownames(mtcars)

mtcars_tbl_hp <- mtcars_tbl %>% select(name, hp)
colnames(mtcars_tbl_hp)
colnames(mtcars_tbl_hp) <- c("car", "horsepower") # assign new column names
mtcars_tbl_hp
colnames(mtcars_tbl_hp)[1] <- "carname"  # assign new column name by index
mtcars_tbl_hp
```

### data.frame indexing
```{r, df4}
mtcars[1, 1] # value of one cell, order is row then column
mtcars[1, ] # row to new data.frame
mtcars[, 1] # column to vector
mtcars[, -c(1:5)] # negative selection
mtcars[c(1:2), "hp"] # combination of number index and names
```

## 3. matrix, combination of multiple vectors with the same types

### matrix vs df
```{r, matrix1}
mtcars_mat <- as.matrix(mtcars)
mtcars_mat 
mtcars_tbl_mat <- as.matrix(mtcars_tbl)
mtcars_tbl_mat # all coerced to character

object.size(mtcars)
object.size(mtcars_mat) # smaller and faster with certain calculations
```

### very similar function calls to df
```{r, matrix2}
# mtcars_mat$mpg # can't do this for matrix
# mtcars_mat[["mpg"]] # can't do this for matrix

dim(mtcars_mat)
ncol(mtcars_mat)
nrow(mtcars_mat)
colnames(mtcars_mat)
rownames(mtcars_mat)

length(mtcars) # number of cols
length(mtcars_mat) # number of cells, probably want to avoid using them

colnames(mtcars_mat)[1] <- "milespergallon" # assign new column names
mtcars_mat
```

### matrix indexing
```{r, matrix3}
mtcars_mat[1, ] # row to vector, named vector
mtcars_mat[, 1] # column to vector
mtcars_mat[1, 1] # value of one cell
mtcars_mat[, -c(1:5)] # negative selection
mtcars_mat[c(1:2), "hp"] # combination of number index and names

t(mtcars_mat)
```

### mathematical operations
```{r, matrix4}
log2(mtcars_mat)
rowSums(mtcars_mat)
rowMeans(mtcars_mat)
colSums(mtcars_mat)
colMeans(mtcars_mat)

mtcars_mat - 5 # -5 on every numeric value
mtcars_mat * c(1, 0, -1) # vector recycling, note sequence
mtcars_mat - mtcars_mat[, 1] # each element in the same row is subtracted by the corresponding vector element, ie normalize by the 1st column
```

### matrix practice question -
How would you center the data (subtract the mean for each variable/column)? 
```{r, matrix_prac}
#hint: start with 
mtcars_mat_t <- t(mtcars_mat)
```

## 4. list, collection of objects(vectors, matrices, data.frames, etc)
### lists store all types of data
```{r, list1}
l1 <- list(1, 
           c("what", "ever"), 
           mtcars)
l1

l2 <- list(n = 1, 
           c = c("what", "ever"), 
           df = mtcars, 
           l = l1) # with names, also can even include lists
l2

pbda::cc.genes # example of list
```

### exploring list
```{r, list2}
cc.genes[[1]]
cc.genes$s.genes

length(cc.genes)
length(cc.genes[[1]])
names(cc.genes)

unlist(cc.genes) # named vector
unlist(cc.genes, use.names = FALSE)

c(cc.genes, "geneA") # combine into list
cc.genes[[1]] <- c(cc.genes[[1]], "geneA") # combine into first list element
```

### list practice question - 
From `cc.genes`, how many markers are shared for both S phase and G2/M?
```{r, list_prac}
```

### list practice question 2 - 
Are ggplot objects list?
```{r, list_prac2}
library(ggplot2)
g <- ggplot(mtcars, aes(x = hp, y = mpg, color = factor(cyl))) + geom_point()
```

## 5. S3/S4, more complex objects, similar to lists but accessed with $ and @. See RNAseq class.

## 6. factors, grouping variables the data
### plotting use case
```{r}
library(ggplot2)
months_tbl <- data.frame(month = c("Jan", "Feb", "Mar"),
                         labmeetings = c(0, 3, 9))
ggplot(months_tbl, aes(x = month, y = labmeetings)) +
  geom_col() +
  cowplot::theme_cowplot() # <- ordered alphabetical, not ideal

months_tbl_factor <- months_tbl %>% mutate(month = factor(month, levels = c("Jan", "Feb", "Mar")))
months_tbl_factor$month
ggplot(months_tbl_factor, aes(x = month, y = labmeetings)) +
  geom_col() +
  cowplot::theme_cowplot() # <- ordered as described by levels
```

### for other uses of factors, please visit https://forcats.tidyverse.org/

# I/O, reading and writing files

## readr package

### faster more flexible than base R functions for large files
```{r, io_readr}
library(readr) 
path <- system.file("extdata", "gene_tibble.csv", package = 'pbda') # data included in package
# call `less` from the terminal, paste in path as well
gene_tbl <- read_csv(path)
write_csv(gene_tbl, "gene_tbl.csv")
write_csv(gene_tbl, "gene_tbl.csv.gz") # will auto zip if indicated
getwd() # saved here if full path is not given

path2 <- system.file("extdata", "hg19_genes.bed.gz", package = 'pbda') # will auto unzip
bed_tbl <- read_tsv(path2) # use col_names = FALSE or give vector of names
# use the terminal and zless to look at it briefly

david_tbl <- read_tsv("https://raw.githubusercontent.com/IDPT7810/practical-data-analysis/master/inst/extdata/david.txt") # link directly

write_lines(cc.genes$s.genes, "Sgenes.txt") # write vector into file, each element on a line

?read_delim # worth looking through the options
```

### we recommend using readr functions, with "_", but beware of the rowname exclusion
```{r, readr2}
write_csv(mtcars, "mtcars.txt")
read_csv("mtcars.txt") # row names are gone!

mtcars %>% as_tibble(rownames = "rowname") # before saving, use one of these two options
mtcars %>% tibble::rownames_to_column("rowname")

write.csv(mtcars, "mtcars2.txt")
r1 <- read_csv("mtcars2.txt") 
r2 <- read.csv("mtcars2.txt") # note the different default column name assignment
r3 <- read.csv("mtcars2.txt", row.names = 1) # gets back row names
```

### readr practice question -
look at this variant call format, read with readr.
rename column names to c("chromosome", "position", "variantID", "ref_allele", "alt_allele", "quality", "filter", "info")
count number of variants at each reported position, sort by descending order
```{r, readr_prac}
# download.file("https://raw.githubusercontent.com/IDPT7810/practical-data-analysis/master/inst/extdata/clinvar_2000.vcf",
#              "clinvar_2000.vcf")
```

## readxl for xlsx files
```{r, readxl}
library(readxl)
read_excel(readxl_example("datasets.xlsx"))
?read_excel
```

## saving objects as .rdata (.rda is the same)

rstudio asks or by default saves entire environment into .rdata file

this causes slow start up and other conflicts, we recommend turning the option off

but do use rdata when needed
```{r, io_rdata}
save(mtcars_mat, mtcars_mat_t, file = "mat.RData")
rm(mtcars_mat, mtcars_mat_t) # a way to remove from environment
load("mat.RData") # load them back
```

## .rds, saving one object, but allows for name reassignment
```{r, io_rds}
saveRDS(mtcars_mat, "mtcars_mat.rds") # default compress = TRUE
mat2 <- readRDS("mtcars_mat.rds") 
identical(mat2, mtcars_mat)
```
