---
title: "Class 3: Common Analysis Challenges"
author: "Jay Hesselberth"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Class-3}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "img/class-3-",
  fig.align = "center"
)
```

```{r load}
library(tidyverse)
library(eda)
```

## Missing data

### Zeroes, `NA`, `NaN` and `NULL`

+ Don't use use zeroes to represent missing data. `0` is valid observed value.

+ `NA` (Not Available) is most often use to represent missing data.

+ `NaN` (Not a Number) is the result of an undefined operation, e.g. `0 / 0`.

+ `NULL` means "undefined" and is only used in a programming context (i.e., a function that returns `NULL`). You can't put `NULL` values in a data frame.

Let's examine a data frame with some missing data.

```{r complete}
# devtools::install_github('rnabioco/eda')
missing
```

### `filter` with `is.na()`

You can identify variables with `NA` values by combining `filter()` and `is.na()`.

```{r filter_na}
# find rows where name is NA
missing %>% filter(is.na(name))

# find rows where id is *not* NA
missing %>% filter(!is.na(id))
```

### `na.omit()`

You can remove **all** rows containing `NA` values with `na.omit()`.

```{r na.omit}
missing %>% na.omit()
```

### `na.rm` argument to stats

Exclude `NA` values from statistical operations with `na.rm = TRUE`.

```{r stat_na}
missing$value1

# if NAs are present, the result is NA
sum(missing$value1)

# solution: drop NAs from the calculation
sum(missing$value1, na.rm = TRUE)
```

### Exercises

1. How many rows of `missing` have `NA` values for *either* `value1` or `value2` (27)?

2. How many rows of `missing` remain after filtering out `id` values of `NA` (25)?

3. Calculate the mean for values in each group in `missing`. Use `summarize()`, then try `summarize_at()`.

## Too much data

### Examining sorted data

```{r head_tail}
sorted <- brauer_gene_exp %>%
  select(systematic_name, rate, expression) %>%
  arrange(rate, desc(expression))

sorted %>% head(10)

sorted %>% tail(10)
```

You can also use the `View()` function to examine the data in a spreadsheet-like viewer.

```{r view, eval=FALSE}
sorted %>% View()
```

### Sampling rows

Use `sample_n()` and `sample_frac()` to randomly select rows from a large data set. Get a reproducible sample using `set.seed()`

```{r sample}
brauer_gene_exp

# sample 1e5 rows
set.seed(47681)
brauer_gene_exp %>% sample_n(1e3)

# sample 10% of the rows
set.seed(47681)
brauer_gene_exp %>% sample_frac(0.1)
```

### Joining tables

Many biological databases store information in multiple tables. These tables are related by a common identifier: a gene name, or possibly a numeric identifer.

We will use two datasets `brauer_gene_exp` and `yeast_prot_prop` to illustrate how data from related tables can be combined using joining operations.

- `brauer_gene_exp` contains gene expression data from yeast grown under different nutrient limitation conditions.

- `yeast_prot_prop` contains descriptive data for yeast proteins.

Examine the two tables and identify common variables for linking. If the variables have different names, Use the `by` argument to `left_join()` to specify common variables.

```{r left_join}
gene_exp_rate <- brauer_gene_exp %>% select(systematic_name, rate, expression)
gene_mw <- yeast_prot_prop %>% select(ORF, Mw)

# Find rows from gene_exp_rate with a match in gene_mw and return all variables
left_join(gene_exp_rate, gene_mw, by = c("systematic_name" = "ORF"))

# Find rows from gene_exp_rate with a match in gene_mw and return only the gene_exp_rate variables
semi_join(gene_exp_rate, gene_mw, by = c("systematic_name" = "ORF"))

# Find rows from gene_exp_rate *without* a match in gene_mw
anti_join(gene_exp_rate, gene_mw, by = c("systematic_name" = "ORF"))
```

### Exercises

1. Calucate mean expression values for each `systematic_name` in `brauer_gene_exp`. How do the numbers change after sampling 1% of the data?

2. How many unique gene names (`systematic_name`) are shared between the `brauer_gene_exp` and `yeast_prot_prop` tables? (5,536).

## Too many variables

We'll use the `brauer_gene_exp` data set to illustrate how you can use summary statistics to capture relationships between many variables.

This plot illustrates the relationship between gene expression and growth rate for the Grr1/YJR090C gene under glucose limitation.

```{r brauer_plot}
library(cowplot)

exp_data <- brauer_gene_exp %>%
  filter(nutrient == "Glucose" & systematic_name == "YJR090C")

exp_data %>%
  ggplot(aes(x = rate, y = expression)) +
  geom_point() +
  geom_smooth(method = 'lm', se = FALSE) + 
  ggtitle("Gene expression vs. growth rate for Grr3 under glucose limitation")
```

You can also use `lm()` to model this relationship.

```{r brauer_lm}
lm(expression ~ rate, exp_data)
```

The results of many statistical operations (`lm()`, `t.test()`, `cor.test()`) are not tidy and relevant parameters are not easily assessible. The `broom` package provides a function called `tidy()` that tidies the results of these operations. 

```{r brauer_tidy}
library(broom)
fit <- exp_data %>% lm(expression ~ rate, data = .)
tidy(fit)
```

### Summaries for mutliple groups

Can we capture linear relationships between expression and rate for all combinations of variables? This is surprisingly easy.

```{r brauer_mult_models, eval = FALSE}
models <- brauer_gene_exp %>%
  group_by(systematic_name, name, nutrient) %>%
  do(model = tidy(lm(expression ~ rate, data = .)))
```

Plot the expression / rate relationship for the top-ten most significant genes. Color lines by nutrient.

```{r brauer_plot_models, eval = FALSE}
topten <- models %>%
  unnest() %>%
  filter(name != "") %>%
  arrange(p.value) %>%
  head(10) %>%
  select(systematic_name)

topten_exp <- brauer_gene_exp %>% semi_join(topten)

topten_exp %>%
  ggplot(aes(x = rate, y = expression, color = nutrient)) +
  geom_point() +
  geom_smooth(method = 'lm', se = FALSE) +
  facet_wrap(~systematic_name + name) +
  scale_color_brewer(palette = "Set1")
```

### Exercises

1. 

