---
title: "Class 3: Common Analysis Challenges"
author: "Jay Hesselberth"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Class-3}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "img/class-3-",
  fig.align = "center"
)
```

```{r load}
library(tidyverse)
library(eda)
```

## Missing data

### Zeroes, `NA`, and `NULL`

Don't use use zeroes to represent missing data. `0` is an observed value 

`NA` (Not Available) is most often use to represent missing data.

`NULL` means "undefined" and is only used in a programming context (i.e., a function that returns `NULL`). You can't put `NULL` values in a data frame.

Let's examine a data frame with some missing data.

```{r complete}
# devtools::install_github('rnabioco/eda')
missing
```

### `filter` with `is.na()`

You can identify variables with `NA` values by combining `filter()` and `is.na()`.

```{r filter_na}
# find rows where name is NA
missing %>% filter(is.na(name))

# find rows where id is *not* NA
missing %>% filter(!is.na(id))
```

### `na.omit()`

You can remove all rows containing `NA` values with `na.omit()`.

```{r na.omit}
missing %>% na.omit()
```

### `na.rm` argument to stats

You can exclude `NA` values from statistical operations with `na.rm = TRUE`.

```{r stat_na}
# if NAs are present, the result is NA
sum(missing$value1)

# solution: drop NAs from the calculation
sum(missing$value1, na.rm = TRUE)
```

## Too much data

### Databases

### Sampling rows

+ `head()` and `tail()`

+ `sample()` and `sample_n()`

+ `set.seed()`

### Reprex

```{r reprex}
```

## Too many variables (identifying important ones, and summarizing others. using colors and facets in plots)

- Assessing significance (light statistics overview, null distributions, hypothesis testing).

## Assessing significance between groups

### Tests - `t.test()` and `ks.test()`

### Tidying results with `broom`

