---
title: "Exercises 2"
author: "Your Name Here"
date: "12/3/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

Here is a [link](https://raw.githubusercontent.com/rnabioco/eda/master/vignettes/exercises-2.Rmd) to the text of these exercises.

# Question 1

Calculate the mean and variance of the `Ozone`, `Solar.R`, `Wind`, and `Temp` variables for each month in the `airquality` dataset, ignoring `NA` values. Use `summarize()` first, then use `summarize_at()`. Note that arguments to summary functions can be passed as dots to `summarize_at()`.

## Strategy

```{r q1}

```

## Interpretation

# Question 2

Calculate mean expression values for each `systematic_name` and `nutrient` combination of `brauer_gene_exp`. Calculate the values by sampling 1%, 10% and 50% of the rows. How do your numbers change at each threshold?

## Strategy

```{r q2}

```

## Interpretation

# Question 3 

Join the `brauer_gene_exp` and `yeast_prot_prop` tables and calculate the mean codon adaptation index (`CAI` in `yeast_prot_prop`) for each molecular funtion (`MF` in `brauer_gene_exp`). Sort the results from largest to smallest mean CAI.

## Strategy

```{r q3}

```

## Interpretation
