---
title: "problem-set-2"
output: html_document
---

# Question 1

Write a function that plots rate vs expression when provided a data.frame, a systematic_name, and a nutrient. Add the common gene name as a title and the nutrient as a subtitle. Include an argument that lets the user select whether a regression line is shown.
```{r, fig.width = 6, fig.height = 3.5}
plot_expr <- function(input, sys_name, nutri, add_lm = F) {

  # Finish writing the function
  
}

plot_expr(
  input = brauer_gene_exp, 
  sys_name = "YDL104C",
  nutri = "Ammonia",
  add_lm = T
)
```

# Question 2

Add an `if` statement to check whether the systematic_name and nutrient provided by the user are present in the input data.
```{r, fig.width = 6, fig.height = 3.5}
plot_expr <- function(input, sys_name, nutri, add_lm = F) {

  # Finish writing the function
  
}

plot_expr(
  input = brauer_gene_exp, 
  sys_name = "YDL104C",
  nutri = "Ammonia",
  add_lm = T
)
```

# Question 3

Add the ellipsis so the user can pass arguments directly to `theme()`.
```{r, fig.width = 6, fig.height = 3.5}
plot_expr <- function(input, sys_name, nutri, add_lm = F) {

  # Finish writing the function
  
}

plot_expr(
  input = brauer_gene_exp, 
  sys_name = "YDL104C",
  nutri = "Ammonia",
  add_lm = T,
  axis.title = element_text(size = 20),
  axis.text = element_text(face = "italic")
)
```

# Question 4

Use your function with `map()` to create a figure containing plots for four genes
```{r, fig.width = 10, fig.height = 6.5}
sys_names <- c("YNL049C", "YML017W", "YDL104C", "YLR115W")

expr_plots <- map(

  # Add the arguments for map()
  
)

plot_grid(plotlist = expr_plots)
```

